### 2017-03-03 paper/chapters/abstract.tex
last-minute changes

Deep directed generative models leverage the possibilities of deep learning to learn rich, hierarchical structures from datasets in a possibly unsupervised context.
These models can be used to understand high-dimensional, real-world data and transform them into compact representations. Similar to how humans learn to generalize from sensory input, we discuss several models which can learn compact, internal representational structures.
After discussing generative modeling in general and its applications, we then introduce the sigmoid belief network which is a simple generative architecture.
Then we focus on two more recently proposed architectures: the variational autoencoders (VAE), and generative adversarial networks (GAN).
We discuss the key idea of both models and their respective learning and training procedure, including variational inference and adversarial training for VAE and GAN respectively.
Additionally to the originally proposed models, we summarize several extensions and present experimental results for the variational autoencoder.

>>>>

Deep directed generative models leverage the possibilities of deep learning to learn rich, hierarchical structures from datasets in an unsupervised context.
These models can be used to understand high-dimensional, real-world data and transform them into compact representations. Similar to how humans learn to generalize from sensory input, we discuss several models which can learn compact, internal representational structures.
After discussing generative modeling in general and its applications, we introduce the sigmoid belief network which is a simple generative architecture.
We then focus on two more recently proposed architectures: variational autoencoders, and generative adversarial networks.
We discuss the key idea of both models and their respective architecture and training procedure, including variational inference and adversarial training for the variational autoencoder and generative adversarial network respectively.
Additionally to the originally proposed models, we summarize several extensions and present experimental results for the variational autoencoder.

### 2017-03-03 paper/chapters/conclusion.tex
last-minute changes

In this article we have taken a look at different directed models which perform generative modeling using deep neural networks.\\
After a short overview over traditional generative models and models not covered in this article, we have taken a look at a wide area of applications using generative models.\\\\
%
As one of the first basic generative models, the sigmoid belief net lays the foundation for directed generative models and displays its weaknesses during inference and training.\\
The variational autoencoders have been discussed in more detail using variational inference to turn the inference problem into an optimization problem and solving it there.
VAE and its various extensions have been shown to be a flexible tool applicable for different tasks.\\
%
Generative adversarial networks introduced in Section~\ref{sec:gan} uses a game-theoretic approach to generative modeling. In recent years, GANs have been extended of which we discussed a subset.\\\\
%
Deep generative models have shown the capability to learn rich internal representation of high-dimensional data in order to generate samples using recent advances in deep learning.
Going further, as intelligent agents will need to have accurate and generalized internal representations inferred through sensory input to reason about the world, generative models are likely to be the cornerstones of more advanced intelligent systems in the future.

>>>>

In this article, we have taken a look at various directed models which perform generative modeling using deep neural networks.\\
After a short overview over traditional generative models and models not covered in this article, we have taken a look at a wide area of applications using generative models.\\\\
%
As one of the first basic generative models, the sigmoid belief net lays the foundation for directed generative models and displays its weaknesses during inference and training.\\
The variational autoencoders have been discussed in more detail using variational inference to turn the inference problem into an optimization problem and solving it there.
VAE and its various extensions have been shown to be a flexible tool applicable for different tasks.\\
%
Generative adversarial networks introduced in Section~\ref{sec:gan} uses a game-theoretic approach to generative modeling. In recent years, GANs have been extended of which we discussed a subset.\\\\
%
Deep generative models have shown the capability to learn rich internal representation of high-dimensional data in order to generate samples using recent advances in deep learning.
Going further, as intelligent agents will need to have accurate and generalized internal representations inferred through sensory input to reason about the world, generative models are likely to be the cornerstones of more advanced intelligent systems in the future.

### 2017-03-03 paper/chapters/gan.tex
last-minute changes

%%% when introduced, game theory approach
Generative Adversarial Networks (GAN) is a framework proposed in 2014 by Goodfellow et al. \cite{gan:2014} based on a game theoretic approach to generative modeling~\cite[Chapter~20.10.4]{deeplearning:2016}.
%which has gathered a lot of attention in the deep learning community and in the last few years many extensions have been [proposed] (citations).

GANs circumvent the issues arising around the inference problem by playing a two-player game. Both players play non-cooperatively against each other, which means they only try to maximize their respective success.
One player is called the generator $G$ which tries to produce authentic looking data samples while the other player, the discrimininator $D$ has to decide whether a received sample is generated or actually real data.
It is important to note that the generator gets to know whether the discriminator accepted the generated output as \emph{real} or rejected it as \emph{fake}.

@@

\paragraph{Solution}
The game-theoretic solution to the emerging situation where neither $G$ nor $D$ can improve its own success with respect to the outcome of the game gets called \emph{Nash equilibrium}\cite{game_theory:1994}.
In the case of GANs, this situation arises when the generator produces samples which get rejected with probability $p=0.5$ by the discriminator. In the case of such equilibrium, the generator produces good enough samples to fool the generator in about half cases.\\\\
\todo[inline]{Add some text showing the theoretical foundation of GANs}

Another formulation of the same game is to define the zero-sum game between the generative model and its adversary, the discriminator~\cite[Chapter~20.10.4]{deeplearning:2016}.
The payoff of the discriminator can be described as the value function $V(G,D)$ while the payoff of the generator is the negative payoff of $D$: $-V(G,D)$.\\
This formulation allows us to define the reached nash equlibium in equation \ref{eq:zerosum_convergence}.
\begin{equation}
  \label{eq:zerosum_convergence}
  G^* = \arg \min_{G} \max_{D} V(G,D)
\end{equation}

@@

However, for gradient-based learning $\mathcal{L}_g$ provides poor gradient during learning \todo[inline]{early in training! show difference.} where the discriminator is able to reject samples from the generator with high confidence.\\
Therefore is equation \ref{eq:l_g_2} more commonly used in practice \todo[inline]{citations! dcgan? improved gan?}.

@@

\begin{figure}[htb]
\centering
  \includegraphics[width=\linewidth]{media/gan_training-crop}
  \caption[Learning Procedure in a GAN]{Learning a GAN on a one-dimensional distribution from the original paper\cite{gan:2014}}
  \label{fig:gan_training}
  \medskip
  \small
  The line named $z$ denotes the noise prior, while the arrows to the line indicated by $x$ is the transformation learned by the generator network. Three different distributions are showed, the solid line represents the generated sample in each step, while the dots are sampled from the data distribution $p_{data}(x)$. The dashed line shows the decision boundary with which probability the discriminator can distinguish between real and fake data.\\
  (a) Prior to training when both $D$ and $G$ are poor and $p_{model}$ does not match $p_{data}$.\\
  (b) After a step of learning $D$ with fixed $G$, the discriminator is able to distinguish fake from real samples easily as indicated by the decision boundary.\\
  (c) After a step of learning $G$ with fixed $D$ has the generator learned to produce samples more similar to $p_{data}(x)$.\\
  (d) The model has reached the nash equilibrium where $p_{model} = p_{data}$ and the discriminator can only distinguish between real and fake sample with $p=0.5$.
\end{figure}

Combining the two objectives into one value function yields equation \ref{eq:minimax_value} which is used to define minmax-game played by the two networks.
\begin{equation}
\label{eq:minimax_value}
\min_G \max_D V(D,G) = \mathbb{E}_{x \sim p_{data}}[\log D(x)] + \mathbb{E}_{z \sim p_z(z)}[\log(1 - D(G(z)))]
\end{equation}

@@

%\paragraph{Theoretical Foundation}
Using a model with infinite training time as well as infinite capacity in both generator and discriminator allows to formulate the nash equilibrium in the generative adversarial network.
For any given generator $G$ and data points $x \sim p_{data}(x)$, the optimal discriminator $D$ is given by equation \ref{eq:opt_d}.
\begin{equation}
  \label{eq:opt_d}
  D_G^*(x) = \frac{p_{data}(x)}{p_{data}(x) + p_{model}(x)}
\end{equation}

@@

\subsubsection{Freeze Learning}
\label{ssub:gan_freeze_learning}
One of the main issues prominent during GAN training is the disparity in performance between $G$ and $D$.
Most of the time during training will one network outperform the other player by such great margin that the gradients from the stronger player will vanish resulting in even slower training for the weaker network.
Thus one of the common practices\cite{improved_gan:2016}\cite{dcgan:2015} for learning GANs is to freeze one network and prevent it from learning until the other player has catched up.
%In practice the generator is far slower to train than the discriminator in most cases (citation).

\subsubsection{Label Smoothing \cite{improved_gan:2016}}
\label{ssub:gan_label_smoothing}
Another rather basic idea to improve gradient flow is to replace the binary classification from the discriminator with smoothed values $0.1/0.9$ or $\epsilon/1-\epsilon$.
This helps the generator with an informative gradient to improve learning performance.
Additionally, smoothed labels have been shown to reduce the attack surface of neural networks to adversarial examples \cite{adv_examples:2016}.
% TODO replace 0/1 with e/1 − e ??
% TODO one-sided label smoothing

@@

\subsubsection{Feature Matching}
\label{ssub:gan_feature_matching}
In computer vision, features are often used to identify relevant pieces of information in images for recognition and classification.
With the recent advances in classification tasks performed by deep convolutional networks, automatically learned features have been a large part of that success.
Feature matching is an technique designed to improve the learning procedure of GANs where $\mathcal{L}_g$ is not not directly maximized but where the discriminator also learns on intermediate layer activation outputs.
Using this modified procedure aids the generator to output data matching the statistics of the input data more closely.
%With $f(x)$ denoting an intermediate layer in $D$, the new objective function becomes:
%$$
%|| \mathbb{E}_{x \sim p_{data}}f(x) - \mathbb{E}_{z \sim p(z)}f(G(z)) || ^{2}_{2}
%$$

@@

\subsubsection{Deep Convolutional GAN}
\label{ssub:dcgan}
Deep convolutional networks\cite{lecun-89e} have yielded impressive results for image classification \cite{imagenet:2012} and have arguably been one of the main reasons for the recent success of deep learning.
These techniques however have mostly been applied in a supervised learning context, not with unsupervised learning.

The discriminator in the GAN model implements a binary classification task and the performance of the generator is heavily dependent on the discriminator, as the training procedure uses the discriminator in the objective function $\mathcal{L}_{g}$ in equation \ref{eq:l_g_1} and \ref{eq:l_g_2}.\\
This leads to the suggestion that using deep convolutional networks helps boost the model performance. In fact is the deep convolution GAN proposed by Radford et al.\cite{dcgan:2015} one of the first GAN variants which results in visually pleasing results.

In order to use convolutional networks in the generator and discriminator, fractionally strided convolutions have been used to basically upsample the image. The exact details of the architecture is beyond the scope of this article, however in addition to the impressive generative performance the authors have shown that the model has learned to produce representional structures in latent space.
Furthermore, as shown in Figure \ref{fig:dcgan_vector} they were able to perform vector artithmetic in the latent space with results indicating that their model was able to arrange the latent space in a useful way.
%\todo[inline]{Elaborate on how DCGAN arranges latent space}

>>>>

%%% when introduced, game theory approach
Generative Adversarial Networks (GAN) is a framework proposed in 2014 by Goodfellow et al. \cite{gan:2014} based on a game-theoretic approach to generative modeling~\cite[Chapter~20.10.4]{deeplearning:2016}.
%which has gathered a lot of attention in the deep learning community and in the last few years many extensions have been [proposed] (citations).

GANs circumvent the issues arising around the inference problem by playing a two-player game. Both players play non-cooperatively against each other, which means they only try to maximize their respective success.
One player is called the generator $G$ which tries to produce authentic looking data samples while the other player, the discrimininator $D$ has to decide whether a received sample is generated or actually real data.
It is important to note that the generator gets to know whether the discriminator accepted the generated output as real or rejected it as fake.

@@

\paragraph{Solution}
The game-theoretic solution to the emerging situation where neither $G$ nor $D$ can improve its own success with respect to the outcome of the game gets called Nash equilibrium\cite{game_theory:1994}.
In the case of GANs, this situation arises when the generator produces samples which get rejected with probability $p=0.5$ by the discriminator. In the case of such equilibrium, the generator produces good enough samples to fool the generator in about half cases.\\\\

Another formulation of the same game is to define the zero-sum game between the generative model and its adversary, the discriminator~\cite[Chapter~20.10.4]{deeplearning:2016}.
The payoff of the discriminator can be described as the value function $V(G,D)$ while the payoff of the generator is the negative payoff of $D$: $-V(G,D)$.\\
This formulation allows us to define the reached nash equilibrium in equation \ref{eq:zerosum_convergence}.
\begin{equation}
  \label{eq:zerosum_convergence}
  G^* = \arg \min_{G} \max_{D} V(G,D)
\end{equation}

@@

However, for gradient-based learning $\mathcal{L}_g$ provides poor gradient during learning where the discriminator is able to reject samples from the generator with high confidence.
Therefore is Equation \ref{eq:l_g_2} advised to be used instead of Equation \ref{eq:l_g_1}~\cite{gan:2014}.\\

@@

\begin{figure}[htb]
\centering
  \includegraphics[width=\linewidth]{media/gan_training-crop}
  \caption[Learning Procedure in a GAN]{Learning a GAN on a one-dimensional distribution from the original paper\cite{gan:2014}}
  \label{fig:gan_training}
  \medskip
  \small
  The line named $z$ denotes the noise prior, while the arrows to the line indicated by $x$ is the transformation learned by the generator network. Three different distributions are showed, the solid line represents the generated sample in each step, while the dots are sampled from the data distribution $p_{data}(x)$. The dashed line shows the decision boundary with which probability the discriminator can distinguish between real and fake data.\\
  (a) Prior to training when both $D$ and $G$ are poor and $p_{model}$ does not match $p_{data}$.\\
  (b) After a step of learning $D$ with fixed $G$, the discriminator is able to distinguish fake from real samples easily as indicated by the decision boundary.\\
  (c) After a step of learning $G$ with fixed $D$ has the generator learned to produce samples more similar to $p_{data}(x)$.\\
  (d) The model has reached the nash equilibrium where $p_{model} = p_{data}$ and the discriminator can only distinguish between real and fake samples with $p=0.5$.
\end{figure}

Combining the two objectives into one value function yields equation \ref{eq:minimax_value} which is used to define the minmax-game played by the two networks.
\begin{equation}
\label{eq:minimax_value}
\min_G \max_D V(D,G) = \mathbb{E}_{x \sim p_{data}}[\log D(x)] + \mathbb{E}_{z \sim p_z(z)}[\log(1 - D(G(z)))]
\end{equation}

@@

%\paragraph{Theoretical Foundation}
Using a model with infinite training time as well as infinite capacity in both generator and discriminator allows formulating the nash equilibrium in the generative adversarial network.
For any given generator $G$ and data points $x \sim p_{data}(x)$, the optimal discriminator $D$ is given by equation \ref{eq:opt_d}.
\begin{equation}
  \label{eq:opt_d}
  D_G^*(x) = \frac{p_{data}(x)}{p_{data}(x) + p_{model}(x)}
\end{equation}

@@

\subsubsection{Freeze Learning}
\label{ssub:gan_freeze_learning}
One of the main issues prominent during GAN training is the disparity in performance between $G$ and $D$.
Most of the time during training will one network outperform the other player by such great margin that the gradients from the stronger player will vanish resulting in even slower training for the weaker network.
Thus one of the common practices\cite{improved_gan:2016}\cite{dcgan:2015} for learning GANs is to freeze one network and prevent it from learning until the other player has caught up.
%In practice the generator is far slower to train than the discriminator in most cases (citation).

\subsubsection{Label Smoothing}
\label{ssub:gan_label_smoothing}
Another rather basic idea to improve gradient flow is to replace the binary classification from the discriminator with smoothed values $0.1/0.9$ or $\epsilon/1-\epsilon$\cite{improved_gan:2016}.
This helps the generator with an informative gradient to improve learning performance.
Additionally, smoothed labels have been shown to reduce the attack surface of neural networks to adversarial examples \cite{adv_examples:2016}.
% TODO replace 0/1 with e/1 − e ??
% TODO one-sided label smoothing

@@

\subsubsection{Feature Matching}
\label{ssub:gan_feature_matching}
In computer vision, features are often used to identify relevant pieces of information in images for recognition and classification.
With the recent advances in classification tasks performed by deep convolutional networks, automatically learned features have been a large part of that success.
Feature matching is an technique designed to improve the learning procedure of GANs where $\mathcal{L}_g$ is not directly maximized but where the discriminator also learns on intermediate layer activation outputs.
Using this modified procedure aids the generator to output data matching the statistics of the input data more closely.
%With $f(x)$ denoting an intermediate layer in $D$, the new objective function becomes:
%$$
%|| \mathbb{E}_{x \sim p_{data}}f(x) - \mathbb{E}_{z \sim p(z)}f(G(z)) || ^{2}_{2}
%$$

@@

\subsubsection{Deep Convolutional GAN}
\label{ssub:dcgan}
Deep convolutional networks\cite{lecun-89e} have yielded impressive results for image classification \cite{imagenet:2012} and have arguably been one of the main reasons for the recent success of deep learning.

The discriminator in the GAN model implements a binary classification task and the performance of the generator is heavily dependent on the discriminator, as the training procedure uses the discriminator in the objective function $\mathcal{L}_{g}$ in equation \ref{eq:l_g_1} and \ref{eq:l_g_2}.\\
This leads to the suggestion that using deep convolutional networks helps boost the model performance. In fact is the deep convolution GAN proposed by Radford et al.\cite{dcgan:2015} one of the first GAN variants which result in visually pleasing results.

In order to use convolutional networks in the generator and discriminator, fractionally strided convolutions (also called deconvolution~\cite{deconv:2010}) have been used to basically upsample the image. The exact details of the architecture is beyond the scope of this article, however in addition to the impressive generative performance the authors have shown that the model has learned to produce representational structures in latent space.
Furthermore, as shown in Figure \ref{fig:dcgan_vector} they were able to perform vector arithmetic in the latent space with results indicating that their model was able to arrange the latent space in a useful way.
%\todo[inline]{Elaborate on how DCGAN arranges latent space}

### 2017-03-03 paper/chapters/introduction.tex
last-minute changes

"What I cannot create, I do not understand." -- Richard Feynman \\\\

Directed generative models are a wide class of models used for generating samples from an unknown distribution $p_{data}$.
%Lets denote $p_{data}$ as the unknown real-world data distribution from which samples in the given dataset are drawn.
The presented models in this article will approximate $p_{data}$ with $p_{model}$ which enables us to draw samples from the model.
Under the assumption that $p_{model}$ approximates $p_{data}$ accurately, samples from the true data distribution should have high likelihood under the approximation as well.\\\\

\paragraph{Short-term applications}
\begin{itemize}
  \item \textbf{Data augmentation} is a technique to enrich an existing dataset by adding more derived information to provide more in-depth insight.
  \item \textbf{Semi-supervised learning} on datasets with a small amount of unlabelled data and a large amount of labeled data to improve model performance. Examples include the conditional variational autoencoder discussed in \ref{ssub:cvae}.
  \item \textbf{Data exploration and understanding} is often needed especially for high-dimensional and complex data. By using models which reduce the dimensions by selecting only important features for generative modeling can help to improve data understanding. Figure \ref{fig:vae_experiments} explores the two-dimensional manifold of a handwritten digit, which originally had 784 dimensions.
\end{itemize}

In addition to those short-term applications, there exist possibilities for more long-term aspects of generative modeling.
For example with unsupervised representation learning, where statistical models are able to learn robust, accurate and generalized internal representations from high-dimensional sensory input can help power future agents to help understand real-world data, allowing these agents to take actions and reason about the real world.

\newpage
%Deep learning techniques have provided great success in applying gradient-based optimization methods to deep neural networks to achieve impressive results in discriminative models.
%These models require however large amounts of labelled data, which is often expensive or infeasible to obtain.
%Generative models can be used to learn \emph{representational} structures from rich, unlabelled data by having less parameters than the input data and are thus forced to capture only essential information.\\
%\newpage

%Prior to discussion individual models, we need to declare a few assumptions and definitions.
%Additionally we assume that samples $x$ are also conditionally dependent on some hidden structure $z$: $x \sim p(x|z)$. 
%\begin{equation}
  %z \sim p(z)\\
  %x \sim p(x|z)
%\end{equation}

%\paragraph{Sampling}
%means for any given $z$ to draw a sample $x$ from our model.
%Most commonly, sampling is the easiest requirement to fulfil.

%\paragraph{Learning}
%of the model parameters through maximum likelihood estimation generally involves $p(x)$ or its gradient.\\
%To get $p(x)$, we first need to marginalize our the latent variables.
%\begin{equation}
  %p(x) = \sum_h p(x|z)p(z)
%\end{equation}
%Unfortunately, in the case of deep learning with many layers this is intractable, as we need to evaluate all model configurations.
%via maximum likelihood generally involves the gradients of $p(x)$
%\paragraph{Inference}
% motivation / what we want exactly and why
%Building intelligent systems requires having accurate and generalized information about the environment in order to perform actions or reason about the world.
%Inferring structured and meaning information from unlabelled data, a process called unsupervised learning, remains a challenging task.

%However, due to the massive increases in availability of data emerging from the web as well as the ongoing progress in computational power there has been significant progress using this data.

%Generative modelling aims to generate new data similar to previously seen input, in a way which is sound with the internal structure of the data.

%% on building reasonable representations
%The assumption that data can be explained using an underlying, simpler structure than the raw input data is important to handle the high-dimensional nature of most data source.
Models using deep learning often assume that observed data $x$ is stochastically dependent on some hidden structure $z$.
To understand how these statistical models help us to discover the data structure, we will take a look at a few examples of high-dimensional real-world data.

\begin{itemize}
  \item \textbf{Images} consisting of millions of pixels can be interpreted as data vectors with each pixel representing an own dimension.
  However, these dimensions are highly correlated and structured, for example nearby pixels have with high probability similar values.
  Looking at images of handwritten digits, the data can be reduced to a handful of variables which may include the digit number, stroke width, position or size.
  Statistical models help us discover these relations and reduce them to representations in lower-dimensional space.
  %\item commonly have thousands of dimensions, but the underlying structure is often far simpler. For example are 28x28 images of handwritten digits 784-dimensional but assuming one-hot encoding the most important information - the digit - it can be explained using 10 dimensions. There are a few more dimensions, stroke width, cursive, position, size. But still far lower than 784. We will take a look at this example later on in \ref{sec:vae}.
  \item \textbf{Audio data} is omni-present in the real world, however most data sources can be transformed from raw audio samples to spoken words or musical notes without losing too much information for reconstruction.
  %\item \textbf{...}
\end{itemize}
% cope with
%To develop systems learning from real world data we use statistical models which are able to infer the underlying structure from
%Real world data is often high-dimensional and noise and transforming this data into an compact representation is therefore an important and also challenging task to perform.
%Generative models are ideal for this task since it has to understand the underlying structure to generate similar data.
%Building this structure is called \textbf{representation learning} and is key to many generative models and how they perform.\\

%With this assumption in mind we can take a look at some examples of input data.

The assumption that the observed variables $x$ can be explained using the causal factors $z$ can be expressed as graphical model shown in Figure \ref{fig:dgm}.
%as having the conditional probability $p(x|z)$ where the latent variables $z$ are the causal factors. This stochastic dependency can be represented in a graphical model shown in figure \ref{fig:dgm}.\\

\begin{figure}[htb]
\centering
\includegraphics{media/directed_graphical_model}
  \caption{Directed graphical model}
  \label{fig:dgm}
  \medskip
  \small
  The latent variable $z$ is a stochastic dependency of the observed data $x$.
\end{figure}

\newpage
%Inverting $p(x|z)$ yields $p(z|x)$ which is one of the main tasks of some generative models, for example the VAE \ref{sec:vae}.

%After a short overview of different approaches and methods and an introductory example with the sigmoid belief net (SBN), we will take a closer look at two recent promising frameworks, the variational autoencoder (VAE \ref{sec:vae}) and the generative adversarial network (GAN \ref{sec:gan}).

%Generative models are one of the most promising ways to perform unsupervised learning (citation needed).
%More specifically, generative modelling assumes a hidden structure $z$ explaining the input data $x$.
%This assumption is reasonable given the following examples:\\

%Generative models have been proposed as one of the most promising approaches towards
%learning representations of real world data by some of the leading researchers (YannLeCun, OpenAI blogpost).

%In recent years supervised learning has yielded impressive results,
%however for these models to succeed huge amount of labelled data is needed.
%Unsupervised learning, that is learning from unlabelled data, has been expressed
%as one hurdles toward general articial intelligence (citation needed).
%Early approaches toward this problem however have shown that these problems are very hard to learn (intractable?).

%In generative models the approach is different in a way that
%we search for a good internal representation of the data.
%This resembles the way humans learn about the world where we incrementally build
%a model of the world.

>>>>

%%%
%%% intuition // generative vs discriminative
%%%
Intuitively, generative modeling is based on the idea that in order to produce data similar to samples from a dataset it helps to understand the data first~\cite[p.~720]{deeplearning:2016}.
%The case of deep learning suggests the use of deep, hierarchical architectures to capture dependencies within the data and reduce the high-dimensional input~\cite{reducing_dim:2006}.
In contrast to discriminative models which model the conditional distribution $p(y|x)$ of the labels $y$ given observations $x$, generative algorithms model the joint distribution $p(x,y)$ over both labels and observations to describe how the data was generated~\cite{learning_dbn:2006}.
As there are only observations given, the generative task seems more difficult than classification where both input and desired output are given during training.~\cite[p.~695]{deeplearning:2016}\\\\
%
Due to the nature of generative models, they are unsupervised learning algorithms which mean that they try to learn the structure of samples from an unlabeled dataset\cite[p.~105-106]{deeplearning:2016}\cite{prob_unsupervised:1999}.
%
Generative models in machine learning using probabilistic models can be mostly separated into directed and undirected models~\cite[p.~77]{deeplearning:2016}.
%
This separation indicates how the model can be mapped to a graph with random variables as vertices and interactions between random variables as edges.
Directed models induce a directed network graph while undirected models induce a network graph with undirected edges.\\\\
%
%
In directed models, we assume that the observations $x$ are stochastically dependent on the underlying factors, also called latent variables $z$. This relationship can be expressed as the graphical model shown in Figure~\ref{fig:dgm}.
Note that the latent variables can be structured hierarchically in order to express different levels of granularity in the features of the underlying factors and representations~\cite[p.~557]{deeplearning:2016}.
\begin{figure}[htb]
\centering
  \includegraphics[width=1.3cm]{media/directed_graphical_model}
  \caption[Graphical Model]{Directed Graphical Model}
\label{fig:dgm}
  \medskip
  \small
  The observed data $x$ is stochastically dependent on the latent variables $z$.\\
  Based on~\cite[Chapter 8]{bishop:2006}.
\end{figure}

\subsection{Overview}
% traditional models (HMM, GMM)
Traditional models include hidden Markov models used for example in speech recognition~\cite{hmm:1989}, which are generative models~\cite[p.~878]{modern_approach:2009} and Gaussian mixture models~\cite[p.~190]{deeplearning:2016}\cite{schuster:1999}.\\\\
%
% undirected networks (Boltzmann machine)
Undirected generative models include various extensions of the Boltzmann machine~\cite{bm:1985} such as the restricted Boltzmann machine (RBM), first described as Harmonium~\cite{smolensky:1986} and the deep Boltzmann machine~\cite{dbm:2009}\cite[Chapter~20.4]{deeplearning:2016}.\\\\
%
% directed networks (SBN, DBN)
For directed models, the sigmoid belief network (SBN) is an early model described in 1992~\cite{neal:1992} which is presented in Section~\ref{sec:sbn}.
SBN has also been extended with a restricted Boltzmann machine for the first two layers resulting in deep belief nets (DBN)~\cite{learning_dbn:2006}, which have been attributed to spark the beginning of current deep learning~\cite[p.~662]{deeplearning:2016}. DBN are partially directed models, as they contain both directed and undirected connections~\cite[p.~664]{deeplearning:2016}.\\\\
%
%Another class of generative models are auto-regressive networks (ARN), which 
%\newpage
%%%
%%% applications
%%%

\subsection{Applications}

\begin{figure}[t]
  \includegraphics[width=\linewidth]{media/caption2img}
  \caption[Caption-to-Image Samples]{Generative Text to Image Synthesis~\cite{gan_t2i:2016}}
  \medskip
  \small
  The top row shows the provided captions and in the bottom row the corresponding synthesized images.
\label{fig:caption2img}
\end{figure}

Generative models learn the joint distribution $p(x,y)$ and due to this fact they can be used to perform classification as well by expand $p(x,y)$ to $p(y|x)$ as indicated by Equation~\ref{eq:class}.
\begin{equation}
  \label{eq:class}
  p(y|x) = \frac{p(x,y)}{p(x)}
\end{equation}
%
One obvious application of generative models is to use the model to sample more data similar to samples from a given training dataset.\\
%
In computer vision tasks, generative models have been largely outperformed by much simpler models until at least 2011~\cite{thesis:2011}. In 2015 however, convolutional networks have been shown to surpass those models in image synthesis~\cite{synthesis:2015}.
Other tasks involving image modeling include performing super-resolution~\cite{superres:2017}, constructing 3D models from 2D images~\cite{gan_3d:2016}, predicting the next video frames~\cite{gan_video:2016}, and synthesizing new images from captions as shown in Figure~\ref{fig:caption2img}.\\
%
In the area of reinforcement learning, generative models offer a way to perform action space exploration~\cite{rl_variational:2016}. Generative adversarial networks discussed in Section~\ref{sec:gan} have been shown to be used in imitation learning, where the generative model is used to imitate previously seen interactions~\cite{imitation_learning:2016}.\\
%
SampleRNN~\cite{samplernn:2016}, which is a recurrent neural network combined with an auto-regressive network have been shown to produce good generated audio samples.\\\\
%
Recently, generative models have also shown to improve semi-supervised learning performance significantly~\cite{cvae:2014}. Semi-supervised learning algorithms use only small amount of labeled data, which is usually difficult to obtain and large amounts of unlabeled data, which is easier to obtain in most cases.

\newpage

### 2017-03-03 paper/chapters/sigmoid_belief_nets.tex
last-minute changes

\section{Sigmoid Belief Networks}
\label{sec:sbn}
\begin{figure}[htb]
\centering
\input{media/sbn_architecture}
  \caption[Architecture of Sigmoid Belief Nets]{Conceptual architecture of belief nets}
\label{fig:sbn_arch}
  \medskip
  \small
  Bayesian networks, also called belief nets are an acyclic directed graph where blue nodes indicate hidden variables and red nodes represent observed variables.
  Arrows indicate stochastic dependency.
  Even though most belief nets are divided into many layers in the context of deep learning, the original definition does not specify that and thus indicates the dashed arrow the possibility of inter-layer dependencies.
\end{figure}

Sigmoid Belief Networks (SBN) were described in 1992 by R.M. Neal~\cite{neal:1992} representing a specific type of bayesian networks~\cite{pearl:1985} shown in Figure~\ref{fig:sbn_arch}
In the case of SBNs however, all states are binary and the activation function is the sigmoid function. They are one of the earliest neural networks used for generative modeling, predating all other presented models in this article.
%As described in the overview above, SBNs contain a number of binary states $s$, most commonly divided into many layers.
Each state $s_i$ is a random variable in the bayesian framework which means it can be observed, hidden or unknown representing different states of knowledge. As such it can also be interpreted with a probability distribution.
\begin{equation}
  \label{eq:sbn_node}
p(s_i) = \sigma\bigg(\sum_{j<i}W_{j,i}s_j+b_i\bigg)
\end{equation}
where 
\begin{conditions}
  \sigma(x) & $\frac{1}{1 + e^{-x}}$ (sigmoid function)\\
  W_{j,i} & connection weights from state $s_j$ to $s_i$\\
\end{conditions}

To draw samples from this network, it is possible to evaluate Equation~\ref{eq:sbn_node} for each visible node.
However prior to that, the SBN first needs to learn how to generate good samples which includes the \emph{learning} and the \emph{inference} problem.\\
Learning means in this context to adjust the weights and biases of the network to be more likely to generate data similar to samples from the given training dataset.
Meanwhile statistical inference is tasked to infer the states of the hidden nodes based on the states in the visible layer.\\

Due to each state being statistically depend on all ancestors, inference means to evaluate all possible configurations of every previous node.
Especially in deep networks, this kind of computation is infeasible as it requires computing exponentially many configurations.

>>>>

\section{Sigmoid Belief Networks}
\label{sec:sbn}
\begin{figure}[htb]
\centering
\input{media/sbn_architecture}
  \caption[Architecture of Sigmoid Belief Nets]{Conceptual architecture of belief nets}
\label{fig:sbn_arch}
  \medskip
  \small
  Bayesian networks, also called belief nets are an acyclic directed graph where blue nodes indicate hidden variables and red nodes represent observed variables.
  Even though most belief nets are divided into many layers in the context of deep learning~\cite[Chapter 20.10.1]{deeplearning:2016}, the original definition does not specify that and thus indicates the dashed arrow the possibility of inter-layer dependencies.
\end{figure}

Sigmoid Belief Networks (SBN) were described in 1992 by R.M. Neal~\cite{neal:1992} representing a specific type of bayesian networks~\cite{pearl:1985} shown in Figure~\ref{fig:sbn_arch}
In the case of SBNs however, all states are binary and the activation function is the sigmoid function. They are one of the earliest neural networks used for generative modeling, predating all other presented directed models in this article.
%As described in the overview above, SBNs contain a number of binary states $s$, most commonly divided into many layers.
Each state $s_i$ is a random variable in the bayesian framework which means it can be observed, hidden or unknown representing different states of knowledge. As such it can also be interpreted with a probability distribution.
\begin{equation}
  \label{eq:sbn_node}
p(s_i) = \sigma\bigg(\sum_{j<i}W_{j,i}s_j+b_i\bigg)
\end{equation}
where
\begin{conditions}
  \sigma(x) & $\frac{1}{1 + e^{-x}}$ (sigmoid function)\\
  W_{j,i} & connection weights from state $s_j$ to $s_i$\\
\end{conditions}

To draw samples from this network, it is possible to evaluate Equation~\ref{eq:sbn_node} for each visible node.
However, prior to that the SBN first needs to learn how to generate good samples which includes the \emph{learning} and the \emph{inference} problem~\cite[p.~695]{deeplearning:2016}.\\
Learning means in this context to adjust the weights and biases of the network to be more likely to generate data similar to samples from the given training dataset.
Meanwhile, statistical inference is tasked to infer the states of the hidden nodes based on the states in the visible layer.\\
%
Due to each state being statistically dependent on all ancestors, inference means to evaluate all possible configurations of every previous node.
Especially in deep networks, this kind of computation is infeasible as it requires computing exponentially many configurations.
